New and simplified manual controls for projection and slice tours, with application to exploring classification boundaries in high dimensions

### 2023-02-07 paper/paper.tex
Getting ready to re-submit

From a statistical perspective 3D is a rare data dimension, so unlike in
most 3D rotation computer graphics applications,
\textcolor{blue}{methods which work for arbitrary dimensions have more general utility}.
A good approach is to show projections from an arbitrary dimensional
space to create dynamic data visualizations called \emph{tours}. Tours
involve views of high-dimensional data with low-dimensional projections.
In his original paper on the grand tour, \citet{As85} provided several
algorithms for tour paths that could theoretically show the viewer the
data \emph{from all sides}. (Various videos of tours, e.g.
\citet{dataviewer}, \citet{gt-pp-video}, on the \citet{ASA22},
illustrate this concept.) Prior to Asimov's work, there were numerous
preparatory developments including \citet{tukey}'s PRIM-9. PRIM-9 had
user-controlled rotations on coordinate axes, allowing one to manually
tour through low-dimensional projections. (A video \citep{PRIM9-video}
illustrating the capabilities is available through \textcolor{blue}{the}
video library of \citet{ASA22}.) Steering through all possible
projections is impossible, unlike Asimov's tours which allow one to
quickly see many, many different projections. After Asimov, there have
been many tour developments, which are summarized in \citet{lee2021}.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Provide \(A\), and \(m\). (Note that \(m\) could also be automatically
  chosen as the component that is closest to the cursor position.)
\item
  Change values in row \(m\), for example, if \(d=2\) gives \[
  A^* = [ \boldsymbol{a}^*_1~\boldsymbol{a}^*_2 ] = \left[ \begin{array}{cc} a_{11} & a_{12}\\
                             \vdots & \vdots \\
                             a^*_{m1} & a^*_{m2}\\
                             \vdots & \vdots \\
                             a_{p1} & a_{p2} 
       \end{array}\right].
  \] \noindent A large change in these values would correspond to making
  a large jump from the current projection. Small changes would
  correspond to tracking a cursor, making small jumps from the current
  projection.
\item
  Orthonormalise \(A^*\), using Gram-Schmidt.

@@

@@

\textcolor{blue}{Within R, the natural implementation for an interactive interface would use}
\texttt{shiny} \citep{shinym} with \texttt{plotly} \citep{plotly}
drawing. However, the \texttt{plotly} package is designed to add to
\texttt{ggplot2} graphics, not to be a programmable interactive graphics
system. It lacks the capability to implement the primary tool needed for
a manual tour, which is the tracking of the cursor position in small
increments. In theory, the historical \texttt{locator()} function of
base R should provide the necessary capability, but its behavior is not
consistent between graphical devices. The cursor position control is
possible with \texttt{rgl} \citep{rgl} but this package is focused on
3D, not the arbitrary dimension of tours. The level of interactivity
would have been possible with the \texttt{cranvas} system
\citep{cranvas} or \texttt{rJava} \citep{rJava} but neither has a
graphics toolkit that currently operates across platforms. What we have
chosen to do is to use the interactivity of Mathematica to experiment,
and to implement the resulting ideas as an animation in the
\texttt{tourr} \citep{tourr} package. There is hope that better
interactivity will be available for R soon, perhaps in javascript
\citep[e.g.][\citet{detourr}]{langevitour}, which might allow for the
experimentation to develop new techniques.

@@

The examples presented in the application section below (and attached in
notebook format in the supplementary material) all use our primary new
function, \texttt{SliceDynamic}. This function typically accepts grouped
data in the form of a matrix where the second last column details the
name of the group and the last column details the group index (which can
be one if there is only one group). After specifying the initial slice
thickness and the slice range, the user is presented with an interactive
display in which the control objects appear on the left and the slice
visualization appears on the right. The user can change the orientation
of the slice via the locator pane, which changes the projection matrix;
a slider controls the slice thickness; and there is an input field,
which changes the slice center. The user can also change the appearance
of the plot by zooming into the center or changing the point size with
the sliders provided. It is worth noting that this zoom works best with
data that has been centered and scaled. The projected data can be
displayed to contrast it with the slice by ticking a box. The explicit
projection matrix can also be displayed via another checkbox. This is
especially useful when we wish to import a projection that was
identified as interesting in the manual exploration into a later study,
as shown in the applications below. More details about the
implementation and usage instructions are given in the Appendix.

@@

\caption{Scatterplot matrix of the (standardized) penguins data. The three species are reasonably different in size, with Gentoo distinguished from the other two on body depth relative to flipper length and body mass.}\label{fig:penguins-scatmat}
\end{figure}

@@

We start with a comparison of the two models and the data distribution
shifting to \(\mathbf{c}^{+}\), thus the slice is localized towards high
values of bill depth in the top row of Figure \ref{slice1p}.
\textcolor{blue}{The star glyph at right indicates the slice center value, which shows that it is higher on `bd` and at the middle value for the other variables. This slice center guide is as proposed in @condviz2, and more details can be found in the Appendix.}
We can see that all three slices (the two models and the data) contain
almost no points from the third class (Gentoo) and that the decision
boundary between the two models is very similar.

@@

\begin{itemize}
\tightlist
\item
  The Mathematica source code defining the new functions in mmtour.wl.
\item
  Three Mathematica notebooks with the code used in the application
  (corresponding to the three subsections).
\item
  Appendix with additional details about manual controls, and
  Mathematica functions.
\item
  The \texttt{Rmd} file that generated the paper, containing the R code
  to produce Figure 1.
\item
  Data files for the examples.
\item
  Animations illustrating the manual tour and slicing, matching the
  static figures in the paper. These are also available at
  \url{https://vimeo.com/747585410} and
  \url{https://vimeo.com/747590472}.
\end{itemize}

>>>>

From a statistical perspective 3D is a rare data dimension, so unlike in
most 3D rotation computer graphics applications,
\textcolor{blue}{methods which work for arbitrary dimensions have more general utility}.
A good approach is to show projections from an arbitrary dimensional
space to create dynamic data visualizations called \emph{tours}. Tours
involve views of high-dimensional data with low-dimensional projections.
In his original paper on the grand tour, \citet{As85} provided several
algorithms for tour paths that could theoretically show the viewer the
data \emph{from all sides}.
\textcolor{blue}{(Various videos of tours, e.g.} \citet{dataviewer},
\citet{gt-pp-video}, on the \citet{ASA22},
\textcolor{blue}{illustrate this concept.)} Prior to Asimov's work,
there were numerous preparatory developments including \citet{tukey}'s
PRIM-9. PRIM-9 had user-controlled rotations on coordinate axes,
allowing one to manually tour through low-dimensional projections. (A
video \citep{PRIM9-video} illustrating the capabilities is available
through \textcolor{blue}{the} video library of \citet{ASA22}.) Steering
through all possible projections is impossible, unlike Asimov's tours
which allow one to quickly see many, many different projections. After
Asimov, there have been many tour developments, which are summarized in
\citet{lee2021}.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Provide \(A\), and \(m\). (Note that \(m\) could also be automatically
  chosen as the component that is closest to the cursor position.)
\item
  Change values in row \(m\), for example, if \(d=2\) \[
  A^* = [ \boldsymbol{a}^*_1~\boldsymbol{a}^*_2 ] = \left[ \begin{array}{cc} a_{11} & a_{12}\\
                             \vdots & \vdots \\
                             a^*_{m1} & a^*_{m2}\\
                             \vdots & \vdots \\
                             a_{p1} & a_{p2} 
       \end{array}\right].
  \] \noindent A large change in these values would correspond to making
  a large jump from the current projection. Small changes would
  correspond to tracking a cursor, making small jumps from the current
  projection.
\item
  Orthonormalise \(A^*\), using Gram-Schmidt.

@@

\textcolor{red}{a2 should be normalised after the second step? Fig 1: this is now showing tracking, with the yellow dot moving in each image?}

@@

\textcolor{blue}{Within R, the natural implementation for an interactive interface would use}
\texttt{shiny} \citep{shinym} with \texttt{plotly} \citep{plotly}
drawing. However, the \texttt{plotly} package is designed to add to
\texttt{ggplot2} \citep{ggplot2} graphics, not to be a programmable
interactive graphics system. It lacks the capability to implement the
primary tool needed for a manual tour, which is the tracking of the
cursor position in small increments. In theory, the historical
\texttt{locator()} function of base R should provide the necessary
capability, but its behavior is not consistent between graphical
devices. The cursor position control is possible with \texttt{rgl}
\citep{rgl} but this package is focused on 3D, not the arbitrary
dimension of tours. The level of interactivity would have been possible
with the \texttt{cranvas} system \citep{cranvas} or \texttt{rJava}
\citep{rJava} but neither has a graphics toolkit that currently operates
across platforms. What we have chosen to do is to use the interactivity
of Mathematica to experiment, and to implement the resulting ideas as an
animation in the \texttt{tourr} \citep{tourr} package. There is hope
that better interactivity will be available for R soon, perhaps in
javascript \citep[e.g.][\citet{detourr}]{langevitour}, which might allow
for the experimentation to develop new techniques.

\textcolor{red}{plotly does 3D graphs that can be dragged as well, maybe need to clarify further? Are there any references that document shortcomings? Maybe this paragraph should be a subsection?}

@@

The examples presented in the application section below (and attached in
notebook format in the supplementary material) all use our primary new
function, \texttt{SliceDynamic}. This function typically accepts grouped
data in the form of a matrix where the second last column details the
name of the group and the last column details the group index (which can
be one if there is only one group). After specifying the initial slice
thickness and the slice range, the user is presented with an interactive
display in which the control objects appear on the left and the slice
visualization appears on the right. The user can change the orientation
of the slice via the locator pane, which changes the projection matrix;
a slider controls the slice thickness; and there is an input field,
which changes the slice center.
\textcolor{red}{insert sentence about new slice position display here.}
The user can also change the appearance of the plot by zooming into the
center or changing the point size with the sliders provided. It is worth
noting that this zoom works best with data that has been centered and
scaled. The projected data can be displayed to contrast it with the
slice by ticking a box. The explicit projection matrix can also be
displayed via another checkbox. This is especially useful when we wish
to import a projection that was identified as interesting in the manual
exploration into a later study, as shown in the applications below. More
details about the implementation and usage instructions are given in the
Appendix.

@@

\caption{Scatterplot matrix of the (standardized) penguins data. The three species are reasonably different in size, with Gentoo distinguished from the other two on bill depth relative to flipper length and body mass.}\label{fig:penguins-scatmat}
\end{figure}

@@

We start with a comparison of the two models and the data distribution
shifting to \(\mathbf{c}^{+}\), thus the slice is localized towards high
values of bill depth in the top row of Figure \ref{slice1p}.
\textcolor{blue}{The star glyph at the right indicates the slice center value, which shows that it is higher on `bd` and at the middle value for the other variables. This slice center guide is as proposed in @condviz2, and more details can be found in the Appendix.}
We can see that all three slices (the two models and the data) contain
almost no points from the third class (Gentoo) and that the decision
boundary between the two models is very similar.

@@

\begin{itemize}
\tightlist
\item
  The Mathematica source code defining the new functions in mmtour.wl.
\item
  Three Mathematica notebooks with the code used in the application
  (corresponding to the three subsections).
\item
  Appendix with additional details about manual controls, and
  Mathematica functions.
\item
  Data files for the examples.
\item
  R code to reproduce the example data, Figures 1 and 2 in the paper and
  Figures 1 and 2 in the appendix.
\item
  Animations illustrating the manual tour and slicing, matching the
  static figures in the paper. These are also available at
  \url{https://vimeo.com/747585410} and
  \url{https://vimeo.com/747590472}.
\end{itemize}
